A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-11-08 appendix.tex
general clean up

\section{One-fluid dispersion relation for the streaming instability}\label{compressible_streaming}
We consider an unstratified disk with $\Phi = \Phi (r)$ so that $\p_zP = \p_z\rho =
0$. The background $\p_rP/\rho$, $\tepsilon$ are constant
input parameters. Then for the Epstein drag law, Eq. \ref{epstein}, we
have $\mathcal{C}=0$ in the background state. 
We Fourier analyze in $r$ and $z$ so that $\p_z\to \ii k_z$ and
$\p_r\to \ii k_x$ when acting on perturbations, and denote $|\bm{k}|^2
\equiv k_x^2 + k_z^2$. We consider large $k_x$ and background
gradients when compared with that of  perturbations. 
%neglect the radial density gradient compared to that of pertuburbations 
%(formally setting $\p_r\rho=0$). 
 This is also done in most local studies of
dusty disks \citep[e.g.][]{youdin07b}.  
The linearized
equations are, after eliminating the azimuthal velocity:

@@

Consider the limit $k_z=0$. Then Eq. \ref{streaming_incompressible}
becomes 
\begin{align}      %streaming_incompressible
\ii \sigma^2 - \frac{\sigma}{\epsilon \tstop}  - \left(\ii \kappa^2 +
  \frac{k_xF_r}{\epsilon}\right) = 0. \label{kz_zero_si}
\end{align} 
Neglecting the quadratic term leads to stability, $\imag{\sigma}<0$, 
as found by \cite{youdin05a} in their full two-fluid analysis as 
$k_z\to0$.  
% However, the real and imaginary parts of Eq. \ref{kz_zero_si}  are
% \begin{align}
% \omega\left(2 s + \frac{1}{\epsilon\tstop}\right) - \frac{k_xF_r}{\epsilon} & =
%                                                                      0, \label{epi_overs1} \\
% \omega^2 - s^2 - \frac{s}{\epsilon\tstop}  - \kappa^2 &= 0, \label{epi_overs2}
% \end{align}
% so the growth rate satisfies 
% \begin{align*}
% \left( \frac{k_xF_r\tstop}{1 + 2 s\epsilon\tstop}\right)^2 - s^2
%   - \frac{s}{\epsilon \tstop} = \kappa^2. %\omega = \frac{k_xF_r\tstop}{1 + 2 s\epsilon\tstop}
% \end{align*} 
However, solving Eq. \ref{kz_zero_si} explicitly assuming $\epsilon s \tstop \ll 1$ would yield 
\begin{align}
  |\omega| &\simeq \left|k_xF_r\right| \tstop \sim \kappa, \\
  s &\simeq \epsilon \tstop \left[\left(k_xF_r\tstop\right)^2 - \kappa^2\right],\label{epi_grow}
\end{align}
Accordingly, given $k_xF_r$, growth is possible if 
\begin{align}
\tstop > \frac{\kappa}{k_x\left|F_r\right|}\equiv t_{\mathrm{s}0}.\label{epi_grow_criterion}
\end{align}
The mode is marginally stable when $\tstop =
t_{\mathrm{s}0}$. Alternatively, for fixed $\tstop$ growth is enabled
by a sufficiently large $k_xF_r$. Unlike the streaming instability,
which is strongly surpressed when $\epsilon=1$, these modes can still
grow at equal dust-to-gas ratio.

These growing epicycles have $|\omega|\gtrsim \kappa$. 
However, these modes are absent in full two-fluid
models \citep{youdin05a}. The discrepency lies in the fact that the
first-order  one-fluid equations are only valid for low-frequency waves with
$|\sigma|\lesssim \Omega$. Thus only 
low-frequency modes should be retained from analyses based on the
first-order one-fluid approximation.

@@

Caution must be taken if the first-order one-fluid equations are used
to simulate dusty gas. 
%One should ensure that the physical dust-gas instabilty
%(which should have low-frequency in order for the first-order one-fluid
%approximation) has a larger growth rate than that of the spurious
%modes, which is given by Eq. \ref{epi_growth}. 
%Fig. \ref{dusty_growth1} suggests that the
%streaming instability would dominate, provided there is sufficient
%resolution to resolve high $K_z$ modes.  
For example, 2D, razor-thin disks would allow the spurious epicycles
to dominate, since in that case the streaming instability cannot
operate. Fig. \ref{dusty_growth1} suggest that in a realistic 3D disks
where a range of $K_z$ is allowed, the streaming instability should
dominate.

>>>>

\section{One-fluid dispersion relation for the streaming instability}\label{compressible_streaming}
We consider an unstratified disk with $\Phi = \Phi (r)$ so that $\p_zP = \p_z\rho =
0$. The background $\p_rP/\rho$, $\tepsilon$ are constant
input parameters. Then for the Epstein drag law, Eq. \ref{epstein}, we
have $\mathcal{C}=0$ in the background state. 
We Fourier analyze in $r$ and $z$ so that $\p_z\to \ii k_z$ and
$\p_r\to \ii k_x$ when acting on perturbations, and denote $|\bm{k}|^2
\equiv k_x^2 + k_z^2$. We consider large $k_x$ and thus neglect background
gradients when compared to that of perturbations.  
%neglect the radial density gradient compared to that of pertuburbations 
%(formally setting $\p_r\rho=0$). 
 This is also done in most local studies of
dusty disks \citep[e.g.][]{youdin07b}.  
The linearized
equations are, after eliminating the azimuthal velocity:

@@

Consider the limit $k_z=0$. Then Eq. \ref{streaming_incompressible}
becomes 
\begin{align}      %streaming_incompressible
\ii \sigma^2 - \frac{\sigma}{\epsilon \tstop}  - \left(\ii \kappa^2 +
  \frac{k_xF_r}{\epsilon}\right) = 0. \label{kz_zero_si}
\end{align} 
Neglecting the quadratic term leads to stability, $\imag{\sigma}<0$, 
as found by \cite{youdin05a} in their full two-fluid analysis. 
% However, the real and imaginary parts of Eq. \ref{kz_zero_si}  are
% \begin{align}
% \omega\left(2 s + \frac{1}{\epsilon\tstop}\right) - \frac{k_xF_r}{\epsilon} & =
%                                                                      0, \label{epi_overs1} \\
% \omega^2 - s^2 - \frac{s}{\epsilon\tstop}  - \kappa^2 &= 0, \label{epi_overs2}
% \end{align}
% so the growth rate satisfies 
% \begin{align*}
% \left( \frac{k_xF_r\tstop}{1 + 2 s\epsilon\tstop}\right)^2 - s^2
%   - \frac{s}{\epsilon \tstop} = \kappa^2. %\omega = \frac{k_xF_r\tstop}{1 + 2 s\epsilon\tstop}
% \end{align*} 
However, solving Eq. \ref{kz_zero_si} explicitly assuming $\epsilon s \tstop \ll 1$ would yield 
\begin{align}
  |\omega| &\simeq \left|k_xF_r\right| \tstop \sim \kappa, \\
  s &\simeq \epsilon \tstop \left[\left(k_xF_r\tstop\right)^2 - \kappa^2\right],\label{epi_grow}
\end{align}
Accordingly, given $k_xF_r$, growth is possible if 
\begin{align}
\tstop > \frac{\kappa}{k_x\left|F_r\right|}\equiv t_{\mathrm{s}0}.\label{epi_grow_criterion}
\end{align}
The mode is marginally stable when $\tstop =
t_{\mathrm{s}0}$. Alternatively, for fixed $\tstop$ growth is enabled
by a sufficiently large $\left|k_xF_r\right|$. Unlike the streaming instability,
which is strongly surpressed when $\epsilon=1$, these modes can still
grow at equal dust-to-gas ratio.

These growing epicycles with $|\omega|\gtrsim \kappa$ are absent in
full two-fluid models \citep{youdin05a}. The discrepency lies in the
fact that the one-fluid equations, to first order in $\tstop$, are
only valid for low-frequency waves with $|\sigma|\lesssim
\Omega$. Thus only low-frequency modes should be retained from
analyses based on Eqs. \ref{masseq}---\ref{tempeq}.

@@

Caution must be taken if the first-order one-fluid equations are used
to simulate dusty gas. 
%One should ensure that the physical dust-gas instabilty
%(which should have low-frequency in order for the first-order one-fluid
%approximation) has a larger growth rate than that of the spurious
%modes, which is given by Eq. \ref{epi_growth}. 
%Fig. \ref{dusty_growth1} suggests that the
%streaming instability would dominate, provided there is sufficient
%resolution to resolve high $K_z$ modes.  
For example, 2D, razor-thin disks would allow the spurious epicycles
to dominate, since in that case the streaming instability cannot
operate. However, Fig. \ref{dusty_growth1} show that in realistic 3D
disks where a range of $K_z$ is allowed, the streaming instability
should dominate.

### 2016-11-08 linear.tex
general clean up

{\bf note: kappa2 accounts for dust effect assuming smallh=0.05}

%We now consider the streaming instability with $k_z\neq 0$. 
We solve the full dispersion relation, Eq. \ref{streaming_dispersion} for selected cases where SI  
growth rates have been 
calculated analytically from full two-fluid disk models and measured 
from direct numerical particle-gas simulations 
\citep[namely][]{youdin07b,bai10b}. These authors use normalized
wavenumbers $K_{x,z} = \eta r k_{x,z}$ where
\begin{align} 
  \eta \equiv -\frac{1}{2\rhog r\OmK^2}\frac{\p P}{\p r} = 
  \frac{1}{2\left(1-\tepsilon\right)}\frac{F_r}{r\OmK^2}, 
\end{align} 
measures the pressure offset of Keplerian rotation. We use
$\eta=0.05c_s/r\OmK$. They also define the stopping time as
$\tau_\mathrm{s}=\tstop/(1-\tepsilon)$.

@@

{\bf note: important to use largrangian pressure pert properly}

We also checked the growth rates $s$ satisfy 
\begin{align} 
  s = \frac{\left|\sigma\right|^2\imag\left(\Delta P
    \Delta\rho^*/\rho\right)}{2\real\left(\sigma\right)\rho\left(\left|\dd
  v_x\right|^2+\left|\dd
  v_z\right|^2\right)}, \label{si_check}
\end{align}
as implied by Eq. \ref{thermal_instability} and
Eq. \ref{pdv}. %, or directly from
%Eq. \ref{streaming_mass}---\ref{streaming_vz}. 
Interestingly, we find that for cases with
$\epsilon>1$, Eq. \ref{si_check} can be satisified with $\Delta
P\simeq \ii\dd v_x \p_rP/\sigma$, i.e. the radial pressure gradient
is responsible for growth. Conversely, for `linB' with $\epsilon
< 1$, one can approximate $\Delta P \simeq \dd P$ in Eq. \ref{si_check}. This
suggests that for SI in gas-dominated disks, the distinction between
Eulerian and Lagragian pressure perturbations is unimportant. 
{\bf but weak growth in this case}
%However
%in that case the gro

@@

\subsection{Comparison to the convective overstability}
%These slowly-growing modes have $|\omega|\simeq \kappa$, so they correspond to
%overstable epicycles. 
The streaming
instability bears some similarity  to the `convective 
overstability' in dust-free, non-isothermal disks 
\citep{klahr14,lyra14,latter16}. In that case, the combination of an 
unstable radial entropy gradient ($N_r^2  \propto \p_rP\p_rS < 0 $) and
finite thermal relaxation time leads to epicyclics with
growth rate $\propto  -N_r^2$

Indeed, the effective radial buoyancy frequency in our dusty disk 
with negligble background density gradient is unstable, 
\begin{align}
  N_r^2 \equiv F_r\frac{\p S_\mathrm{eff}}{\p r} =
  -\frac{\rho}{P}F_r^2<0.  
\end{align}
Convective overstability cannot exist in our locally isothermal disks,
which have zero cooling times. On the other hand, dust-gas drag 
is equivalent to heating/cooling in our framework.

One might then think of the streaming instability as a dusty relative
to the convective  overstability. Explicit solutions in the small
$\tstop$ limit show that streaming instability growth rates are
also $\propto |N_r^2|$  \citep{jacquet11}, as is for the convective
overstability \citep{lyra14}. Furthermore, both instabilities require
finite effective cooling times, and both require $k_z\neq0$. 
{\bf note: si osc freq not too small compared to kappa}

However, the convective overstability is most unstable at $k_x=0$
\citep{latter14}  whereas the streaming instability growth rate is maximized at finite $k_x$
\citep{youdin05a}. Nevertheless, they share the same underlying
physical mechanism, i.e. pressure-density phase lags.

>>>>

%{\bf note: kappa2 accounts for dust effect assuming smallh=0.05}

We solve the full dispersion relation, Eq. \ref{streaming_dispersion} for selected cases where SI  
growth rates have been 
calculated analytically from full two-fluid disk models and measured 
from direct numerical particle-gas simulations 
\citep[namely][]{youdin07b,bai10b}. These authors use normalized
wavenumbers $K_{x,z} = \eta r k_{x,z}$ where
\begin{align} 
  \eta \equiv -\frac{1}{2\rhog r\OmK^2}\frac{\p P}{\p r} = 
  \frac{1}{2\left(1-\tepsilon\right)}\frac{F_r}{r\OmK^2}, 
\end{align} 
measures the pressure offset of Keplerian rotation. We use
$\eta=0.05c_s/r\OmK$. They also define the stopping time as
$\tau_\mathrm{s}=\tstop/(1-\tepsilon)$.

@@

%{\bf note: important to use largrangian pressure pert properly}

We also checked the growth rates $s$ satisfy 
\begin{align} 
  s = \frac{\left|\sigma\right|^2\imag\left(\Delta P
    \Delta\rho^*/\rho\right)}{2\real\left(\sigma\right)\rho\left(\left|\dd
  v_x\right|^2+\left|\dd
  v_z\right|^2\right)}, \label{si_check}
\end{align}
as implied by Eq. \ref{thermal_instability} and
Eq. \ref{pdv}. %, or directly from
%Eq. \ref{streaming_mass}---\ref{streaming_vz}. 
Interestingly, we find that for cases with
$\epsilon>1$, Eq. \ref{si_check} can be satisified with $\Delta
P\simeq \ii\dd v_x \p_rP/\sigma$, i.e. the radial pressure gradient
is responsible for growth. Conversely, for `linB' with $\epsilon
< 1$, one can approximate $\Delta P \simeq \dd P$ in Eq. \ref{si_check}. This
suggests that for SI in gas-dominated disks, the distinction between
Eulerian and Lagragian pressure perturbations is unimportant. 
%{\bf but weak growth in this case}
%However
%in that case the gro

@@

\subsection{Comparison to the convective overstability}
%These slowly-growing modes have $|\omega|\simeq \kappa$, so they correspond to
%overstable epicycles. 
The streaming
instability bears some similarity  to the `convective 
overstability' in dust-free, non-isothermal disks 
\citep{klahr14,lyra14,latter16}. In that case, the combination of an 
unstable radial buoyancy frequency ($N_r^2 < 0 $) and
finite thermal relaxation time leads to growing epicyclics. %with
%growth rate $\propto  -N_r^2$

Indeed, the effective radial buoyancy frequency in our dusty disk 
with negligble background density gradient is unstable, 
\begin{align}
  N_r^2 \equiv -\frac{1}{\rho}\frac{\p P}{\p r}\frac{\p S_\mathrm{eff}}{\p r} =
  -\frac{\rho}{P}F_r^2<0.  
\end{align}
Convective overstability cannot exist in our locally isothermal disks,
which have zero cooling times. On the other hand, dust-gas drag 
is equivalent to heating/cooling in our framework.

One might then think of the streaming instability as a dusty relative
to the convective  overstability. Explicit solutions in the small
$\tstop$ limit show that streaming instability growth rates are
also $\propto |N_r^2|$  \citep{jacquet11}, as is for the convective
overstability \citep{lyra14,latter16}. Furthermore, both instabilities require
finite effective cooling times, and both require $k_z\neq0$. 
%{\bf note: si osc freq not too small compared to kappa}

However, the convective overstability is most unstable at $k_x=0$
\citep{lyra14}  whereas the streaming instability growth rate is maximized at finite $k_x$
\citep{youdin05a}. Nevertheless, they share the same underlying
physical mechanism, i.e. pressure-density phase lags.

### 2016-11-08 model.tex
more expansion on dust-drag numerical section
home/office sync
general clean up

The two fluids interact via a drag force parameterized by a stopping
time $\tstop$ 
such that 
\begin{align}  
  \rhod\left.\frac{\p \bm{v}_\mathrm{d}}{\p t}\right|_\mathrm{drag}= -
  \rhog\left.\frac{\p \bm{v}_\mathrm{g}}{\p t}\right|_\mathrm{drag}=
  - \frac{\rhog\rhod}{\left(\rhod+\rhog\right)}\frac{\left(\bm{v}_\mathrm{d} -
    \bm{v}_\mathrm{g}\right)}{\tstop}. 
\end{align}
is the dust-gas friction force per unit volume. Note that $\tstop$ 
differs slightly from the stopping time $\tau_\mathrm{s} =
\tstop\rhog/\left(\rhog+\rhod\right)$ defined in previous studies
\citep[e.g.][]{youdin05a,jacquet11}.

We consider the limit of tightly-coupled dust particles with 
$\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency. This allows us to simplify the problem by using the
`terminal velocity approximation' so that the dust-gas velocity
difference is 
\begin{align}
  \bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} = \frac{\nabla
    P}{\rhog}\tstop, 
\end{align}
\citep{youdin05a,jacquet11}.

%In the limit of strong dust-gas drag (stopping time $\tstop\to 0$) 
In this limit of strong dust-gas drag, \cite{laibe14} give a single
fluid description of the mixture: 
\begin{align}
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}\\
  &\frac{D T}{D t} = - \left(\gamma-1\right)T\nabla\cdot\bm{v} +
  \mathcal{H}_\mathrm{drag} + \mathcal{H}_\mathrm{eff}  - \Lambda, \label{tempeq} 
\end{align}
where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative \emph{following the mixture} with center-of-mass velocity  
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}, 
\end{align}
and 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total mixture density. For an ideal gas the pressure is given
by $P = \mathcal{R}\rhog T/\mu $, where $\mathcal{R}$ is
the gas constant and $\mu$ is the mean molecular weight.

@@

Nevertheless, since we have shown that an isothermal dusty gas
effectively has $\gamma=1$ we can define an effective (dimensionless)  
entropy for the \emph{entire mixture} as 
\begin{align}
   \seff \equiv \ln \frac{P}{\rho} = \ln{\left[c_s^2(1-\tepsilon)\right]}.  
\end{align} 
We shall see that by defining an effective entropy this way, many of the 
results concerning the stability of dusty gas will have identical form
and interpretations to that in standard adiabatic hydrodynamics of 
pure gas. For example, if $c_s^2$ is constant and $\tstop=0$, then
$D\seff/ D t = 0$ since $D\tepsilon/D t=0$ in that case.

The physical reason for this analogy is that with strong drag, the dust is almost
perfectly entrained in the gas flow, but there is some gain/loss of 
dust particles between different gas parcels if dust-gas
friction if finite. This property is analogous to the entropy of an adiabatic
pure gas subject to heating/cooling: entropy is conserved following
a gas parcel, except if there is heat exchange between a fluid parcel
and the surrounding.

@@

In practice the dust particles settle vertically on a timescale 
$t_\mathrm{settle}\sim 1/\OmK^2\tstop$, although for tightly-coupled dust
particles this is much longer than that to establish vertical
hydrostatic equilibrium in the gas component ($\sim 1/\OmK$).  
True steady states may be calculated by invoking an underlying gas
turbulence, which leads to dust diffusion \citep{takeuchi02, youdin07, 
 lyra13}. This complication is beyond the scope of this paper.

In this work, we will assume that $\tstop$  is sufficiently small so
that settling occurs on much longer timescales  than that of interest; or
consider equilibria with $\mathcal{C}\equiv 0$, namely   
\begin{inparaenum}[1)] 
\item 
  unstratified disks with finite dust-gas drag (\S\ref{si});  
\item 
  perfectly-coupled dust with $\tstop=0$ (\S\ref{results}). 
\end{inparaenum} 
Vertical settling is absent in both cases.

@@

\subsection{Disk structure with a prescribed dust distribution}\label{steady_state}  
To obtain an actual disk structure for numerical computations, we
assume a Gaussian profile in the dust-to-gas ratio,   
\begin{align}\label{dust_gauss}
  \epsilon(r,z) = \epsilon_0(r)
  \exp{\left[-\frac{z^2}{2\Htilde^2(r)}\right]}. 
\end{align}
%For simplicity we assume the dust-to-gas ratio at the mid-plane
%$\epsilon_0$, as well as its characteristic scale-height
%$\widetilde{H}$, are both constant.
% We take the mid-plane dust-to-gas ratio
%to be a power-law in radius,  
%\begin{align}
%  \epsilon_0(r) = \epsilon_{00}\left(\frac{r}{r_0}\right)^{-d},  
%\end{align}
%where $\epsilon_{00}$ is the dust-to-gas ratio at the fiducial radius
%$r_0$.  
Inserting Eq. \ref{dust_gauss} into vertical hydrostatic equilibrium,
Eq. \ref{steady_momz} and integrating with the approximate
gravitational potential (Eq. \ref{thin_disk_potential}) we obtain the
gas density as
\begin{align}
  &\rhog(r,z)= \notag\\
&\rho_\mathrm{g0}(r)\exp{\left\{ - \frac{z^2}{2\Hgas^2}
    -\epsilon_0\frac{\Htilde^2}{\Hgas^2}\left[1 -
      \exp{\left(-\frac{z^2}{2\Htilde^2}\right)}\right] \right\}}, 
\end{align}
where
\begin{align}
  \Hgas = \frac{c_s}{\OmK}, \quad \OmK \equiv \sqrt{\frac{GM_*}{r^3}},   
\end{align}
is the gas scale-height in the dust-free limit and $\OmK$ is the
Keplerian frequency, respectively.

@@

\subsection{Vertical shear}\label{vertshear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p \seff}{\p z} -\frac{\p
    P}{\p z}\frac{\p \seff}{\p r} \right). 
\end{align}
Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align}\label{vshear2}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
%  -\frac{\p\epsilon}{\p z}\left[\frac{\p\ln{\rhog}}{\p r} + \frac{q}{r}\right]\right.\notag\\
 -\frac{\p\epsilon}{\p z}\frac{\p\ln{P}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r} \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z} \right\} 
\end{align}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the third term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient. The last term survives in the dust-free
limit.

@@

{\bf lagrangian derivative follows the mixture. this highlights one
  perk of the one-fluid framework. in two-fluid models there are two
  types of lagragian deriv, not clear which fluid to follow.}

>>>>

The two fluids interact via a drag force parameterized by a stopping
time $\tstop$ 
such that 
\begin{align}  
  \rhod\left.\frac{\p \bm{v}_\mathrm{d}}{\p t}\right|_\mathrm{drag}= -
  \rhog\left.\frac{\p \bm{v}_\mathrm{g}}{\p t}\right|_\mathrm{drag}=
  - \frac{\rhog\rhod}{\left(\rhod+\rhog\right)}\frac{\left(\bm{v}_\mathrm{d} -
    \bm{v}_\mathrm{g}\right)}{\tstop}. 
\end{align}
is the dust-gas friction force per unit volume. Note that $\tstop$ 
differs slightly from the stopping time $\tau_\mathrm{s} =
\tstop\rhog/\left(\rhog+\rhod\right)$ used in some studies
\citep[e.g.][]{youdin05a,jacquet11}.

We consider tightly-coupled dust particles with 
$\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency. This allows us to simplify the problem by using the
`terminal velocity approximation' so that the dust-gas velocity
difference is 
\begin{align}
  \Delta\bm{v}\equiv\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} = \frac{\nabla
    P}{\rhog}\tstop, 
\end{align}
\citep{youdin05a,jacquet11}. More generally, though, $\Delta\bm{v}$
obeys an evolutionary

%In the limit of strong dust-gas drag (stopping time $\tstop\to 0$) 
In this limit of strong dust-gas drag, \cite{laibe14} give a single
fluid description of the mixture, to $O\left(\tstop\right)$, as:  
\begin{align} 
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}\\
  &\frac{D T}{D t} = - \left(\gamma-1\right)T\nabla\cdot\bm{v} +
  \mathcal{H}_\mathrm{drag} + \mathcal{H}_\mathrm{eff}  - \Lambda, \label{tempeq} 
\end{align}
where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative \emph{following the mixture} with center-of-mass velocity  
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}, 
\end{align}
and 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total mixture density. For an ideal gas the pressure is given
by $P = \mathcal{R}\rhog T/\mu $, where $\mathcal{R}$ is
the gas constant and $\mu$ is the mean molecular weight.

@@

Nevertheless, since we have shown that an isothermal dusty gas
effectively has $\gamma=1$ we can define an effective (dimensionless)  
entropy for the \emph{mixture} as 
\begin{align}
   \seff \equiv \ln \frac{P}{\rho} = \ln{\left[c_s^2(1-\tepsilon)\right]}.  
\end{align} 
This follows naturally by combining the effective 
energy Eq. \ref{eff_energy} and Eq. \ref{masseq}. 
With an effective entropy defined this way, many of the 
results concerning the stability of dusty gas will have identical form
and interpretations to that in standard adiabatic hydrodynamics of 
pure gas. For example, if $c_s^2$ is constant and $\tstop=0$, then
$D\seff/ D t = 0$ since $D\tepsilon/D t=0$ in that case.

The physical reason for this analogy is that with strong drag, the
dust is almost perfectly entrained in the gas flow, but there is some
gain/loss of dust particles between different gas parcels. This
property is analogous to the entropy of an adiabatic pure gas subject
to heating/cooling: entropy is conserved following a gas parcel,
except if there is heat exchange between a fluid parcel and the
surrounding.

@@

In practice, dust particles settle to the disk midplane on a timescale 
$t_\mathrm{settle}\sim 1/\OmK^2\tstop$, although for tightly-coupled dust
particles this is much longer than that to establish vertical
hydrostatic equilibrium in the gas component ($\sim 1/\OmK$).  
True steady states may be calculated by invoking an underlying gas
turbulence, which leads to dust diffusion counter-acting vertical
settling \citep{takeuchi02, youdin07}. This complication is beyond the
scope of this paper.

In this work, we assume that $\tstop$  is sufficiently small so
that settling occurs on much longer timescales  than that of interest; or
consider equilibria with $\mathcal{C}\equiv 0$, namely   
\begin{inparaenum}[1)] 
\item 
  unstratified disks with finite dust-gas drag; %(\S\ref{si});  
\item 
  perfectly-coupled dust with $\tstop=0$. %(\S\ref{results}). 
\end{inparaenum} 
Vertical settling is absent in both cases.

@@

\subsection{Disk structure with a prescribed dust distribution}\label{steady_state}  
For numerical computations, we assume a Gaussian profile in the
dust-to-gas ratio,    
\begin{align}\label{dust_gauss}
  \epsilon(r,z) = \epsilon_0(r)
  \exp{\left[-\frac{z^2}{2\Htilde^2(r)}\right]}. 
\end{align}
%For simplicity we assume the dust-to-gas ratio at the mid-plane
%$\epsilon_0$, as well as its characteristic scale-height
%$\widetilde{H}$, are both constant.
% We take the mid-plane dust-to-gas ratio
%to be a power-law in radius,  
%\begin{align}
%  \epsilon_0(r) = \epsilon_{00}\left(\frac{r}{r_0}\right)^{-d},  
%\end{align}
%where $\epsilon_{00}$ is the dust-to-gas ratio at the fiducial radius
%$r_0$.  
Inserting Eq. \ref{dust_gauss} into vertical hydrostatic equilibrium,
Eq. \ref{steady_momz} and integrating with the approximate
gravitational potential (Eq. \ref{thin_disk_potential}) we obtain the
gas density as
\begin{align}
  &\rhog(r,z)= \notag\\
&\rho_\mathrm{g0}(r)\exp{\left\{ - \frac{z^2}{2\Hgas^2}
    -\epsilon_0\frac{\Htilde^2}{\Hgas^2}\left[1 -
      \exp{\left(-\frac{z^2}{2\Htilde^2}\right)}\right] \right\}}, 
\end{align}
where
\begin{align}
  \Hgas = \frac{c_s}{\OmK}, \quad \OmK \equiv \sqrt{\frac{GM_*}{r^3}},   
\end{align}
is the gas scale-height in the dust-free limit and $\OmK$ is the
Keplerian frequency, respectively.

@@

\subsection{Vertical shear}\label{vertshear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p \seff}{\p z} -\frac{\p
    P}{\p z}\frac{\p \seff}{\p r} \right). 
\end{align}
Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align}\label{vshear2}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
%  -\frac{\p\epsilon}{\p z}\left[\frac{\p\ln{\rhog}}{\p r} + \frac{q}{r}\right]\right.\notag\\
 -\frac{\p\epsilon}{\p z}\frac{\p\ln{P}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r} \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z} \right\} 
\end{align}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio. The third term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient, and survives in the dust-free limit.

@@

%{\bf lagrangian derivative follows the mixture.}
% this highlights one
%  perk of the one-fluid framework. in two-fluid models there are two
%  types of lagragian deriv, not clear which fluid to follow.}
